---
title: "Tutorial for R package CTSclocks"
author: "Huige Tong, Xiaolong Guo, Andrew E Teschendorff"
output: 
  BiocStyle::html_document:
    theme: readable
vignette: >
  %\VignetteIndexEntry{Tutorial for R package CTSclocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
knitr::opts_chunk$set(crop = NULL)
```

# Data used in this tutorial

## Bulk PFC data and sorted neuron data

-   The bulk prefrontal cortex (PFC) data is from Murphy et al. This is an Illumina 450k dataset encompassing 38 normal PFC tissue samples and 37 PFC major depressive disorder samples. Only controls are used here. (MurphyGSE88890.rda)
-   The sorted PFC neuron data is from Pai et al. This is an Illumina EPIC dataset encompassing 33 normal neuron tissue samples, 35 schizophrenia samples and 32 bipolar disorder neuron samples. Only controls are used here. (PaiGSE112179.rda)

## A sample data of bulk liver

-   The sample liver data from GSE107038 contains 39 normal liver samples and part of probes. (ExampleData_Liver.Rd)

## Clocks for Brain

-   Neu-In clock (Neu-InCoef.rda)
-   Neu-Ex clock (Neu-ExCoef.rda)
-   Glia-In clock (Glia-InCoef.rda)
-   Glia-Ex clock (Glia-ExCoef.rda)
-   Brain clock (BrainCoef.rda)

## Clocks for Liver

-   Hepatocyte clock (HepClock.rda)
-   Liver clock (LiverClock.rda)

</br></br>

# Application of Neu/Glia/Brain clocks

## Neu-In, Glia-In and Brain clocks

-   For bulk data, these clocks should be applied on centered and scaled to unit standard deviation residuals obtained after adjustment for cell-type fractions.</br>
-   For sorted data, these clocks should be applied on centered and scaled to unit standard deviation beta matrix.</br>
-   Here we apply Neu-In clock on a bulk PFC data from Murphy et al and a sorted neuron data from Pai et al. The Glia-In and Brain clocks work the same way with Neu-In clock.</br> </br>

### Bulk data

```{r message=FALSE}
library(CTSclocks)

data(MurphyGSE88890)
agePred.v = CTSclockAge(beta.m, CTSclock = 'Neu-In', dataType = 'bulk', CTF.m = NULL, tissue = 'brain')
```

```{r echo=FALSE}
library(ggplot2)
phenotype.df$DNAmAgePred = as.numeric(agePred.v)
## Make a scatter plot for DNAm age against chronological age
PvalFormat = function(PvalRaw){
  PvalRaw = as.numeric(PvalRaw)
  if (PvalRaw < 1e-300){
    return('P<1e-300')
  }else{
    if (PvalRaw >= 0.01 & PvalRaw <= 1){
      return(paste0('P=',sprintf("%0.3f", PvalRaw)))
    }else if(PvalRaw < 0.01 & PvalRaw >= 1e-300){
      return(paste0('P=', format(PvalRaw, digits = 1, scientific = T)))
    }
  }
}
p = ggplot(phenotype.df, aes(x = Age, y = DNAmAgePred)) + geom_point(shape = 20, color = 'black') + theme_bw() +
    theme(axis.title.x = element_text(vjust = 2), axis.title.y = element_text(vjust = 2),
          axis.text.x = element_text(vjust = 2), axis.text.y = element_text(vjust = 2),
          plot.title = element_text(hjust = 0.5),
          legend.background = element_rect(linetype = "solid", colour = "black",linewidth = 0.2)) + 
  ggtitle("Neu-In_Murphy(n=38)") + xlab("Chronological Age") + ylab("Predicted Age") + xlim(0, 100) + ylim(0, 100)
corValue = cor.test(phenotype.df$Age, phenotype.df$DNAmAgePred)$estimate
pValue = cor.test(phenotype.df$Age, phenotype.df$DNAmAgePred)$p.value
p = p + annotate("text", x = 20, y = 65, label=paste("PCC=", sprintf("%0.2f", corValue), sep = ""), color = 'darkred', size = 5, fontface = 'italic') + annotate("text", x = 20, y = 85, label = PvalFormat(pValue), color = 'darkred', size = 5, fontface = 'italic')
lm.o = lm(phenotype.df$DNAmAgePred ~ phenotype.df$Age)
a.v = summary(lm.o)$coeff[, 1]
p = p + geom_abline(intercept = 0, slope = 1, color = 'black', linetype = 'dashed', lwd = 0.2) + geom_abline(intercept = a.v[1], slope = a.v[2], color = '#650023', linetype = 'dashed')
print(p)
```

### Sorted data


```{r message=FALSE}
library(CTSclocks)

data(PaiGSE112179)
agePred.v = CTSclockAge(beta.m, CTSclock = 'Neu-In', dataType = 'sorted', CTF.m = NULL, tissue = 'brain')
```

```{r echo=FALSE}
library(ggplot2)
phenotype.df$DNAmAgePred = as.numeric(agePred.v)
## Make a scatter plot for DNAm age against chronological age
PvalFormat = function(PvalRaw){
  PvalRaw = as.numeric(PvalRaw)
  if (PvalRaw < 1e-300){
    return('P<1e-300')
  }else{
    if (PvalRaw >= 0.01 & PvalRaw <= 1){
      return(paste0('P=',sprintf("%0.3f", PvalRaw)))
    }else if(PvalRaw < 0.01 & PvalRaw >= 1e-300){
      return(paste0('P=', format(PvalRaw, digits = 1, scientific = T)))
    }
  }
}
p = ggplot(phenotype.df, aes(x = Age, y = DNAmAgePred)) + geom_point(shape = 20, color = 'black') + theme_bw() +
    theme(axis.title.x = element_text(vjust = 2), axis.title.y = element_text(vjust = 2),
          axis.text.x = element_text(vjust = 2), axis.text.y = element_text(vjust = 2),
          plot.title = element_text(hjust = 0.5),
          legend.background = element_rect(linetype = "solid", colour = "black",linewidth = 0.2)) + 
  ggtitle("Neu-In_Murphy(n=38)") + xlab("Chronological Age") + ylab("Predicted Age") + xlim(0, 100) + ylim(0, 100)
corValue = cor.test(phenotype.df$Age, phenotype.df$DNAmAgePred)$estimate
pValue = cor.test(phenotype.df$Age, phenotype.df$DNAmAgePred)$p.value
p = p + annotate("text", x = 20, y = 65, label=paste("PCC=", sprintf("%0.2f", corValue), sep = ""), color = 'darkred', size = 5, fontface = 'italic') + annotate("text", x = 20, y = 85, label = PvalFormat(pValue), color = 'darkred', size = 5, fontface = 'italic')
lm.o = lm(phenotype.df$DNAmAgePred ~ phenotype.df$Age)
a.v = summary(lm.o)$coeff[, 1]
p = p + geom_abline(intercept = 0, slope = 1, color = 'black', linetype = 'dashed', lwd = 0.2) + geom_abline(intercept = a.v[1], slope = a.v[2], color = '#650023', linetype = 'dashed')
print(p)
```

</br></br>

## Neu-Ex and Glia-Ex clocks

-   For both bulk data and sorted data, these clocks should be applied directly on the beta matrix.
-   Here we apply Neu-In clock on a bulk PFC data from Murphy et al and a sorted neuron data from Pai et al. The Glia-Ex clock works the same way with Neu-Ex clock.

### Bulk data

```{r message=FALSE}
library(CTSclocks)

data(MurphyGSE88890)
agePred.v = CTSclockAge(beta.m, CTSclock = 'Neu-Ex', dataType = 'bulk', CTF.m = NULL, tissue = 'brain')
```

```{r echo=FALSE}
library(ggplot2)
phenotype.df$DNAmAgePred = as.numeric(agePred.v)
## Make a scatter plot for DNAm age against chronological age
PvalFormat = function(PvalRaw){
  PvalRaw = as.numeric(PvalRaw)
  if (PvalRaw < 1e-300){
    return('P<1e-300')
  }else{
    if (PvalRaw >= 0.01 & PvalRaw <= 1){
      return(paste0('P=',sprintf("%0.3f", PvalRaw)))
    }else if(PvalRaw < 0.01 & PvalRaw >= 1e-300){
      return(paste0('P=', format(PvalRaw, digits = 1, scientific = T)))
    }
  }
}
p = ggplot(phenotype.df, aes(x = Age, y = DNAmAgePred)) + geom_point(shape = 20, color = 'black') + theme_bw() +
    theme(axis.title.x = element_text(vjust = 2), axis.title.y = element_text(vjust = 2),
          axis.text.x = element_text(vjust = 2), axis.text.y = element_text(vjust = 2),
          plot.title = element_text(hjust = 0.5),
          legend.background = element_rect(linetype = "solid", colour = "black",linewidth = 0.2)) + 
  ggtitle("Neu-In_Murphy(n=38)") + xlab("Chronological Age") + ylab("Predicted Age") + xlim(0, 100) + ylim(0, 100)
corValue = cor.test(phenotype.df$Age, phenotype.df$DNAmAgePred)$estimate
pValue = cor.test(phenotype.df$Age, phenotype.df$DNAmAgePred)$p.value
p = p + annotate("text", x = 20, y = 65, label=paste("PCC=", sprintf("%0.2f", corValue), sep = ""), color = 'darkred', size = 5, fontface = 'italic') + annotate("text", x = 20, y = 85, label = PvalFormat(pValue), color = 'darkred', size = 5, fontface = 'italic')
lm.o = lm(phenotype.df$DNAmAgePred ~ phenotype.df$Age)
a.v = summary(lm.o)$coeff[, 1]
p = p + geom_abline(intercept = 0, slope = 1, color = 'black', linetype = 'dashed', lwd = 0.2) + geom_abline(intercept = a.v[1], slope = a.v[2], color = '#650023', linetype = 'dashed')
print(p)
```

### Sorted data

```{r message=FALSE}
library(CTSclocks)

data(PaiGSE112179)
agePred.v = CTSclockAge(beta.m, CTSclock = 'Neu-Ex', dataType = 'sorted', CTF.m = NULL, tissue = 'brain')
```

```{r echo=FALSE}
library(ggplot2)
phenotype.df$DNAmAgePred = as.numeric(agePred.v)
## Make a scatter plot for DNAm age against chronological age
PvalFormat = function(PvalRaw){
  PvalRaw = as.numeric(PvalRaw)
  if (PvalRaw < 1e-300){
    return('P<1e-300')
  }else{
    if (PvalRaw >= 0.01 & PvalRaw <= 1){
      return(paste0('P=',sprintf("%0.3f", PvalRaw)))
    }else if(PvalRaw < 0.01 & PvalRaw >= 1e-300){
      return(paste0('P=', format(PvalRaw, digits = 1, scientific = T)))
    }
  }
}
p = ggplot(phenotype.df, aes(x = Age, y = DNAmAgePred)) + geom_point(shape = 20, color = 'black') + theme_bw() +
    theme(axis.title.x = element_text(vjust = 2), axis.title.y = element_text(vjust = 2),
          axis.text.x = element_text(vjust = 2), axis.text.y = element_text(vjust = 2),
          plot.title = element_text(hjust = 0.5),
          legend.background = element_rect(linetype = "solid", colour = "black",linewidth = 0.2)) + 
  ggtitle("Neu-In_Murphy(n=38)") + xlab("Chronological Age") + ylab("Predicted Age") + xlim(0, 100) + ylim(0, 100)
corValue = cor.test(phenotype.df$Age, phenotype.df$DNAmAgePred)$estimate
pValue = cor.test(phenotype.df$Age, phenotype.df$DNAmAgePred)$p.value
p = p + annotate("text", x = 20, y = 65, label=paste("PCC=", sprintf("%0.2f", corValue), sep = ""), color = 'darkred', size = 5, fontface = 'italic') + annotate("text", x = 20, y = 85, label = PvalFormat(pValue), color = 'darkred', size = 5, fontface = 'italic')
lm.o = lm(phenotype.df$DNAmAgePred ~ phenotype.df$Age)
a.v = summary(lm.o)$coeff[, 1]
p = p + geom_abline(intercept = 0, slope = 1, color = 'black', linetype = 'dashed', lwd = 0.2) + geom_abline(intercept = a.v[1], slope = a.v[2], color = '#650023', linetype = 'dashed')
print(p)
```

</br></br>




# Application of Hepatocyte/Liver clocks

**For Hep clock**

```{r message=FALSE}
library(CTSclocks)

data(ExampleData_Liver)
agePred.v = CTSclockAge(Test.m, CTSclock = 'Hep', dataType = 'bulk', CTF.m = NULL, tissue = 'otherTissue')
```

```{r echo=FALSE, message=FALSE}
library(ggpubr)
mae <- median(abs(Age - agePred.v))
# Create a data frame for ggplot2/ggpubr
data <- data.frame(ChronologicalAge = Age, PredictedAge = agePred.v)
ggscatter(
    data, x = "ChronologicalAge", y = "PredictedAge",
    add = "reg.line",                   # Remove automatic regression line
    xlab = "Chronological age (years)",
    ylab = "Predicted age (years)",
    color = "black", 
  ) + 
    geom_abline(slope = 1, intercept = 0, linetype="dashed", color="black") + 
    stat_cor(method = "pearson", label.x = 20, label.y = 90) +
    annotate("text",label=paste0("MedAE: ", format(mae, digits=2), " years"),
              x=30, y=85, size=4) +
    theme_minimal()+
    theme(panel.grid.major = element_blank(),                
          panel.grid.minor = element_blank(),                
          panel.background = element_blank(),                
          axis.line = element_line(),
          axis.ticks=element_line())+
    scale_x_continuous(breaks = seq(10, 90, by = 20), limits = c(10, 90)) +
    scale_y_continuous(breaks = seq(10, 90, by = 20), limits = c(10, 90))

```


**Same steps for the liver clock**

```{r message=FALSE}
library(CTSclocks)

data(ExampleData_Liver)
agePred.v = CTSclockAge(Test.m, CTSclock = 'Liver', dataType = 'bulk', CTF.m = NULL, tissue = 'otherTissue')
```

```{r echo=FALSE, message=FALSE}
library(ggpubr)
mae <- median(abs(Age - agePred.v))
# Create a data frame for ggplot2/ggpubr
data <- data.frame(ChronologicalAge = Age, PredictedAge = agePred.v)
ggscatter(
    data, x = "ChronologicalAge", y = "PredictedAge",
    add = "reg.line",                   # Remove automatic regression line
    xlab = "Chronological age (years)",
    ylab = "Predicted age (years)",
    color = "black", 
  ) + 
    geom_abline(slope = 1, intercept = 0, linetype="dashed", color="black") + 
    stat_cor(method = "pearson", label.x = 20, label.y = 90) +
    annotate("text",label=paste0("MedAE: ", format(mae, digits=2), " years"),
              x=30, y=85, size=4) +
    theme_minimal()+
    theme(panel.grid.major = element_blank(),                
          panel.grid.minor = element_blank(),                
          panel.background = element_blank(),                
          axis.line = element_line(),
          axis.ticks=element_line())+
    scale_x_continuous(breaks = seq(10, 90, by = 20), limits = c(10, 90)) +
    scale_y_continuous(breaks = seq(10, 90, by = 20), limits = c(10, 90))

```
